---
title: "Power & sample-size for the 2 × 2 Social Exclusion experiment"
author: "Ceren"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    embed-resources: true
    theme: cosmo
    fig-width: 9
    fig-height: 6
knitr:
  opts_chunk:
    message: false
    warning: false
    echo: true
    cache: true
execute:
  freeze: auto
---

```{r hidden-setup, include=FALSE}
# Silent setup: install/load packages, set the CRAN mirror (needed for
# non-interactive renders), define the two thresholds we use throughout,
# and pin the two folder paths. Hidden from the rendered document.
options(repos = c(CRAN = "https://cloud.r-project.org"))
required <- c("pwr", "readr", "dplyr", "tidyr", "ggplot2", "knitr")
new <- setdiff(required, rownames(installed.packages()))
if (length(new) > 0) install.packages(new)
invisible(lapply(required, library, character.only = TRUE))

alpha <- 0.05
power <- 0.90

PILOT_DIR  <- "/Users/cerenay/social_exclusion_pilot/processed data and analyses"
DESIGN_DIR <- "/Users/cerenay/social_exclusion_pilot/2025_new_design_and_experiment/Design"
```

## Overview

This document estimates the sample size for the new design. I report
two anchors — our own pilot and Chen & Li(2009)'s identity effects — so we can
compare both before fixing N for the preregistration.

Two complementary approaches:

- **Analytical power** via `pwr::pwr.f2.test()`. Closed-form; returns one
  required N per assumed effect size.
- **Monte Carlo simulation.** A full power curve across sample sizes and
  effect sizes, so the trade-off is visible rather than a single point.

Thresholds throughout: **α = 0.05, power = 0.90**.

## The design being powered

Between-subjects 2 × 2 factorial:

| Factor        | Level 1               | Level 2                                        |
|---------------|-----------------------|------------------------------------------------|
| **Identity** | no identity           | identity (Klee / Kandinsky minimal group)     |
| **Exclusion** | random (computer)     | spectator (independent participant)            |

The four cells correspond to four of the five treatments in the oTree code
(`T1`, `T2`, `T4`, `T5`). `T3` (identity × no exclusion) remains in the
design as a manipulation check but is not part of the 2 × 2 factorial
powered here.

**Primary test:** the Identity × Exclusion **interaction** on dictator-game
giving. Outcome: EMU allocated to the recipient.

## What the pilot tells us

The pilot manipulated a single factor — `condition` (treatment = chosen
exclusion, control = random exclusion) — with no identity component. It
maps onto only one of the new design's two factors (the exclusion
mechanism), so its effect provides a *lower bound* for the interaction of
interest in the new design.


```{r load-pilot}
pilot_treatment <- read_csv2(file.path(PILOT_DIR, "treatment_pilot.csv"))
pilot_control   <- read_csv2(file.path(PILOT_DIR, "control_pilot.csv"))

pilot <- bind_rows(pilot_treatment, pilot_control) |>
  mutate(
    dg   = if_else(role_dg == 1, group_sent, send_proxy),
    fair = case_when(
      payoff_beneficiary == 10 ~ "fair",
      payoff_beneficiary  > 10 ~ "excluder",
      payoff_beneficiary  < 10 ~ "excluded"
    )
  ) |>
  filter(!is.na(dg))

cat("Usable pilot N:", nrow(pilot),
    "  (treatment:", sum(pilot$condition == "treatment"),
    "control:",  sum(pilot$condition == "control"), ")\n")
```

### Marginal means by condition

```{r marginal-means}
pilot |>
  group_by(condition) |>
  summarise(n = n(),
            mean = round(mean(dg), 3),
            sd   = round(sd(dg), 3)) |>
  knitr::kable()
```

### Cell means (condition × fair status)

```{r cell-means}
pilot |>
  group_by(condition, fair) |>
  summarise(n = n(),
            mean = round(mean(dg), 3),
            sd   = round(sd(dg), 3),
            .groups = "drop") |>
  knitr::kable()
```

### Effect sizes we can extract from the pilot

```{r effect-sizes}
d_condition <- with(pilot, {
  tv <- dg[condition == "treatment"]
  cv <- dg[condition == "control"]
  n1 <- length(tv); n2 <- length(cv)
  m1 <- mean(tv);   m2 <- mean(cv)
  s1 <- sd(tv);     s2 <- sd(cv)
  ps <- sqrt(((n1 - 1) * s1^2 + (n2 - 1) * s2^2) / (n1 + n2 - 2))
  (m1 - m2) / ps
})

aov_cond   <- aov(dg ~ condition, data = pilot)
ss         <- summary(aov_cond)[[1]][, "Sum Sq"]
eta2_cond  <- ss[1] / sum(ss)
f_cond     <- sqrt(eta2_cond / (1 - eta2_cond))
pilot_sd   <- sd(pilot$dg)

tibble(
  metric = c("Cohen's d (treatment vs control)",
             "eta^2 (condition)",
             "Cohen's f (condition)",
             "Overall SD of dg (noise anchor for the sims)"),
  value  = round(c(d_condition, eta2_cond, f_cond, pilot_sd), 4)
) |>
  knitr::kable()
```

The pilot's condition effect is **small** on Cohen's benchmarks
(`d ≈ 0.21`, `f ≈ 0.11`, with `f = 0.10` the small threshold). This is
the effect size that drives the pilot-anchored sample sizes reported
below.

## Analytical power under the pilot anchor

The primary test is the Identity × Exclusion interaction in a 2 × 2
between-subjects ANOVA, so:

- Numerator df: `u = (a − 1)(b − 1) = 1`
- Denominator df: `v = N − a·b = N − 4`
- Total N: `v + 4`, cell n: `N / 4`

The table below sweeps four effect-size scenarios so the sample-size can be read directly against the assumed effect size.

```{r analytical-power}
power_2x2_interaction <- function(f, alpha = 0.05, power = 0.90) {
  res <- pwr.f2.test(u = 1, f2 = f^2,
                     sig.level = alpha, power = power)
  N_total <- ceiling(res$v + 4)
  data.frame(
    f          = f,
    f_squared  = round(f^2, 5),
    v_denom_df = ceiling(res$v),
    N_total    = N_total,
    n_per_cell = ceiling(N_total / 4)
  )
}

scenarios <- data.frame(
  scenario = c("Pilot-derived (single-factor)",
               "Small  (Cohen benchmark)",
               "Small-to-medium",
               "Medium (Cohen benchmark)"),
  f        = c(round(f_cond, 3), 0.10, 0.15, 0.25)
)

primary_results <- do.call(rbind, lapply(seq_len(nrow(scenarios)), function(i) {
  cbind(scenario = scenarios$scenario[i],
        power_2x2_interaction(scenarios$f[i], alpha, power))
}))

knitr::kable(primary_results,
             caption = "Sample sizes for the 2 × 2 interaction under the pilot anchor (α = 0.05, power = 0.90)")
```

At the pilot's effect size (`f ≈ 0.11`), ~235 per cell are required for
90 % power on the interaction. At `f = 0.15` (small-to-medium) this drops
to ~120 per cell. At `f = 0.25` (medium) it drops further to ~44 per cell.

## Secondary tests

Two supporting analyses are included.

### 5-treatment omnibus ANOVA

If the analysis strategy is an omnibus F across all five treatments, the required N is slightly different.

```{r five-treatment}
five_treatment <- data.frame(
  scenario = c("Pilot-derived", "Small", "Small-to-medium", "Medium"),
  f        = c(round(f_cond, 3), 0.10, 0.15, 0.25)
)
five_treatment$n_per_cell <- sapply(five_treatment$f, function(f) {
  ceiling(pwr.anova.test(k = 5, f = f,
                         sig.level = alpha, power = power)$n)
})
five_treatment$N_total <- five_treatment$n_per_cell * 5
knitr::kable(five_treatment,
             caption = "Sample sizes for the omnibus 5-treatment ANOVA")
```

### Within-subjects paired contrasts

Under the strategy method each participant supplies paired observations
(Round 2: in-group vs out-group; Round 3: easy vs hard recipient). These
paired tests require far fewer participants than the between-subjects
interaction, so they will be well-powered as a by-product of powering the
interaction. The between-subjects interaction is the binding constraint.

```{r paired-contrasts}
paired_scenarios <- data.frame(
  scenario = c("Small  (d = 0.20)", "Small-to-medium (d = 0.30)",
               "Medium (d = 0.50)"),
  d        = c(0.20, 0.30, 0.50)
)
paired_scenarios$n_needed <- sapply(paired_scenarios$d, function(d) {
  ceiling(pwr.t.test(d = d, sig.level = alpha, power = power,
                     type = "paired", alternative = "two.sided")$n)
})
knitr::kable(paired_scenarios,
             caption = "Sample sizes for within-subjects paired contrasts")
```

## Simulation-based power

For each candidate cell size and each assumed effect size, the routine
below simulates `n_sims` replicate datasets with a crossover cell-mean
pattern (so Cohen's f of the pure interaction equals the target exactly),
fits `aov(y ~ A * B)`, and counts the proportion of simulations in which
the interaction reaches `p < α`. Noise is Normal with SD anchored to the
pilot's DG SD, so residual variance is realistic for the outcome that will
actually be observed.

```{r sim-function}
simulate_power_2x2 <- function(n_per_cell,
                               f,
                               sd     = pilot_sd,
                               n_sims = 500,
                               alpha  = 0.05) {
  delta <- f * sd
  cell_means <- c(-delta, +delta, +delta, -delta)  # A1B1, A1B2, A2B1, A2B2
  A  <- factor(rep(c("A1", "A1", "A2", "A2"), each = n_per_cell))
  B  <- factor(rep(c("B1", "B2", "B1", "B2"), each = n_per_cell))
  mu <- rep(cell_means, each = n_per_cell)
  n_sig <- 0L
  for (i in seq_len(n_sims)) {
    y   <- mu + rnorm(length(mu), sd = sd)
    fit <- aov(y ~ A * B)
    p   <- summary(fit)[[1]]["A:B", "Pr(>F)"]
    if (!is.na(p) && p < alpha) n_sig <- n_sig + 1L
  }
  n_sig / n_sims
}
```

```{r sim-run, cache=TRUE}
set.seed(2026)
n_grid           <- c(20, 40, 60, 80, 100, 120, 150, 200, 250)
f_grid           <- c(0.10, 0.15, 0.20, 0.25)
n_sims_per_point <- 500

sim_grid <- expand.grid(n_per_cell = n_grid, f = f_grid,
                        KEEP.OUT.ATTRS = FALSE)
sim_grid$power <- mapply(
  simulate_power_2x2,
  n_per_cell = sim_grid$n_per_cell,
  f          = sim_grid$f,
  MoreArgs   = list(sd = pilot_sd, n_sims = n_sims_per_point, alpha = alpha)
)
sim_grid$N_total <- sim_grid$n_per_cell * 4
```

### Simulated power table

```{r sim-table}
sim_wide <- pivot_wider(sim_grid[, c("n_per_cell", "f", "power")],
                        names_from = f, values_from = power,
                        names_prefix = "f=")
knitr::kable(sim_wide,
             caption = "Simulated power (rows = cell size, cols = Cohen's f)")
```

### Smallest cell size that reaches 0.90

```{r sim-min}
target_power <- 0.90
achieved <- sim_grid |>
  group_by(f) |>
  filter(power >= target_power) |>
  summarise(n_per_cell_min = min(n_per_cell),
            N_total_min    = min(N_total),
            .groups = "drop")
knitr::kable(achieved,
             caption = sprintf("Smallest n per cell reaching simulated power ≥ %.2f",
                               target_power))
```

### Power curve

The full trade-off surface between sample size and effect size.

```{r sim-plot, fig.cap="Simulated power curves for the 2 × 2 interaction. Dashed lines mark power = 0.80 and 0.90."}
sim_grid$f_label <- paste0("f = ", format(sim_grid$f, nsmall = 2))

plot_power <- ggplot(sim_grid,
                     aes(x = n_per_cell, y = power,
                         color = f_label, group = f_label)) +
  geom_hline(yintercept = c(0.80, 0.90),
             linetype = "dashed", color = "grey55") +
  annotate("text", x = min(sim_grid$n_per_cell), y = 0.80,
           label = "power = 0.80", vjust = -0.4, hjust = 0,
           size = 3.2, color = "grey45") +
  annotate("text", x = min(sim_grid$n_per_cell), y = 0.90,
           label = "power = 0.90", vjust = -0.4, hjust = 0,
           size = 3.2, color = "grey45") +
  geom_line(size = 1) +
  geom_point(size = 2.5) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  scale_x_continuous(breaks = unique(sim_grid$n_per_cell)) +
  labs(
    x     = "Sample size per cell",
    y     = "Simulated power",
    color = "Effect size"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right",
        panel.grid.minor = element_blank())

plot_power
ggsave(file.path(DESIGN_DIR, "power_curves.png"),
       plot = plot_power, width = 9, height = 6, dpi = 150)
```

## Alternative anchor: Chen & Li (2009)

Because our previous pilot is a bit different in terms of identity and exclusion, I also
anchor on Chen & Li (2009, *AER*, 99: 431–457). This is the canonical
Klee/Kandinsky paper — our current design implements their
paradigm directly — so their identity effects are a natural reference for
the *identity* leg of the new design.

### What Chen & Li reported

- **Other-other allocation task.** Ingroup vs outgroup allocation
  difference (normalised by endowment) is **32.2 %–38.4 %** across their
  five rounds, all significant at the 1 % level (Section II results;
  N = 566 subjects total).
- **Identity parameter for charity.** `a = 0.467` — participants weight
  another player's payoff 47 % more heavily when that other player is an
  ingroup match (Table 2 MLE; N = 1896 observations from 432 treatment
  subjects).
- **Identity parameter for envy.** `b = -0.931` — envy toward ingroup
  matches is 93 % lower than toward outgroup matches.
- **Dictator-game choices** (Appendix A, Dict 1–5). Player A ingroup vs
  outgroup choice proportions differ by ~0.12; Player B by ~0.15.

### Translating those to Cohen's f

Two reasonable readings:

- **Behavioural proportion differences.** Cohen's *h* ≈ 0.24–0.32 →
  Cohen's *f* ≈ **0.12–0.16** for the identity main effect.
- **Utility-parameter interpretation.** The 47 % charity increase with a
  ~35 % mean allocation gap → Cohen's *f* ≈ **0.20–0.30**.

One caveat to flag in the preregistration: Chen & Li's numbers describe
their *identity main effect*, not an interaction with a second factor.
The interaction in the new design could plausibly be smaller,
equal, or larger depending on whether identity amplifies the exclusion
mechanism. The four scenarios below span the plausible range.

### Sample sizes under this anchor

```{r chen-li-power}
chen_li_scenarios <- data.frame(
  scenario = c("Chen & Li (behavioural, low)",
               "Chen & Li (behavioural, mid)",
               "Chen & Li (utility param, mid)",
               "Chen & Li (utility param, high)"),
  f        = c(0.13, 0.16, 0.20, 0.25)
)

chen_li_results <- do.call(rbind, lapply(seq_len(nrow(chen_li_scenarios)), function(i) {
  cbind(scenario = chen_li_scenarios$scenario[i],
        power_2x2_interaction(chen_li_scenarios$f[i], alpha, power))
}))
chen_li_results$N_recruit <- ceiling(chen_li_results$N_total / (1 - 0.20))

knitr::kable(
  chen_li_results,
  caption = "Sample sizes under the Chen & Li (2009) anchor (α = 0.05, power = 0.90, 20 % attrition buffer)")
```

### Side-by-side comparison

```{r comparison-side-by-side}
# primary_results doesn't have N_recruit yet (that column gets added below);
# add it here so both frames match before we rbind.
primary_with_buffer <- primary_results
primary_with_buffer$N_recruit <- ceiling(primary_with_buffer$N_total / (1 - 0.20))

cols_for_comparison <- c("scenario", "f", "N_total", "n_per_cell", "N_recruit")
comparison <- rbind(
  cbind(source = "Pilot anchor",     primary_with_buffer[, cols_for_comparison]),
  cbind(source = "Chen & Li anchor", chen_li_results[,      cols_for_comparison])
)
knitr::kable(comparison,
             caption = "Comparison of pilot-anchored and Chen & Li-anchored sample sizes")
```

The Chen & Li range clusters around `f = 0.15–0.20`, which sits between
the pilot-anchored "small" and "medium" scenarios and is materially more
optimistic than the pilot's own marginal (`f = 0.11`). If Chen & Li's
identity effects generalise to the interaction in the new design, the
required sample is ~80–120 per cell rather than ~230–260. The difference
is large enough that both anchors should be reported explicitly in the
preregistration.

## Recommendation

The pilot's own effect (`f ≈ 0.11`) sits at the *small* Cohen threshold.
Powering the 2 × 2 interaction for that effect at 90 % power requires
approximately **1 000 total participants** analytically, and roughly
**1 300 to recruit** after a 20 % attrition buffer. This is expensive on
Prolific.

Chen & Li's identity effects are larger. Depending on the reading,
Cohen's *f* falls in the **0.15–0.25** range. Taking those as the anchor
for the interaction gives **475–600 total** at `f = 0.15`, and
**~175–220 total** at `f = 0.25`.

Given this split, my recommendation for the preregistration:

1. **Primary target: `f = 0.15`** (small-to-medium). ~120 per cell →
   **~475 total → ~590 to recruit** with a 20 % buffer. Defensible on
   both anchors: it lies between the pilot's 0.11 and Chen & Li's low
   end, and matches Chen & Li's behavioural-proportion reading.
2. **`f = 0.25`** as the smallest effect of substantive theoretical
   interest. At that value the design is powered comfortably (~44 per
   cell, ~175 total).
3. **Plan a two-wave / sequential design** if a first wave indicates the
   true effect is closer to `f = 0.10` than to `f = 0.15`.

The between-subjects interaction is the binding constraint. The
within-subjects paired contrasts (in-group vs out-group in Round 2,
easy vs hard recipient in Round 3) will be well-powered as a by-product.

## Attrition / exclusion buffer

Online samples typically lose 10–30 % to attention checks, dropouts, and
comprehension failures. The table below applies a **30 %** buffer to the
pilot-anchored primary results.

```{r buffer}
buffer_rate <- 0.30
primary_results$N_recruit <- ceiling(primary_results$N_total / (1 - buffer_rate))
knitr::kable(
  primary_results[, c("scenario", "f", "N_total", "N_recruit")],
  caption = sprintf("Total N with a %d %% attrition buffer applied",
                    round(100 * buffer_rate)))
```
